\textbf{Political Instability and Financial Markets}

### 2017-08-27 doc/paper.tex
Changed incorrect uses of citep to citet.

A common hypothesis is that political instability discourages investment and reduces economic growth by increasing policy uncertainty. This hypothesis is supported by cross country empirical evidence that political instability - as measured by regime change frequency or political violence - is negatively correlated with investment, GDP growth, and financial development in cross country regressions \citep{alesina1996political,alesina1996income,roe2011political}. \footnote{In particular, \citep{alesina1996political} show that countries have lower growth in years in which there are major government changes such as coups. On the other hand, \citep{londregan1990poverty} finds that coups have no effect on economic growth.} However, such hypotheses are overly simplistic. Not all political instability is equivalent - resignations, assassinations, and coups are each examples of political instability, but may have disparate effects on financial returns. Further, political instability may be seen as a \textit{positive} event if the current regime is regarded as strongly anti-business or anti-global.  \citep{alesina1996political} takes the possibility of a new government adopting better economic policies seriously, but argues that the negative effects of uncertainty dominate the positive effects of coups staged by pro growth factions.

As in \citep{alesina1996political}, we use ``irregular'' regime changes such as coups, assassinations and resignations as an indicator for political instability. The major innovation of this paper is that, in contrast to the aforementioned cross country studies, we analyze the impact of political instability using daily financial data. This enables us to use market expectations to quantify the economic impact of each individual type of regime change. The advantage of this event study approach is that it allows us to isolate the effect of irregular regime changes and mitigate the endogeneity problems that are nearly ubiquitous in cross country regressions.

@@

This paper is closely related to four branches of literature. The first analyzes the economic impact of political instability. This area contains a number of theoretical models in addition to the empirical work already mentioned. These models tend to focus on government policy instruments that affect private rates of return on investment. The general argument is that political instability causes governments to act myopically and adopt inefficient policies.\footnote{For instance, \citep{svensson1998investment} argues that governments are less likely to invest in the legal system and the protection of property rights when political uncertainty is high. Similarly, \citep{devereux1998political} propose a model in which political instability causes governments to leave fewer assets to their successors which forces them into increasing capital taxes. The knowledge of future taxation causes the private sector to reduce current investment which reduces future output.}

The second looks at how the variance of stock market returns reacts to political events. \citep{leblang2005government} show that right-wing governments increase the mean and volatility of stock prices in the United States and Britain. \citep{jensen2005market} find that a rise in the popularity of a Brazilian presidential candidate with uncertain future policies increased the volatility (but not the mean return) of the Brazilian stock market.

The third strand examines the impact of leaders on political and economic outcomes. \citep{jones2005leaders} identify exogenous leadership transitions using deaths of leaders from natural causes or accidents. Their results indicate that leaders influence monetary policy and economic growth. \citep{jones2009hit} find that transitions to democracy are more likely after the assassinations of autocrats.

@@

\section{Impact of Political Instability on Stock Returns} \label{sec: Impact of Political Instability on Stock Returns}
\subsection{Abnormal Returns} \label{subsec: Abnormal Returns}
We begin our analysis by studying the effect of irregular regime changes on stock returns. We follow the standard event study methodology as presented by, among others, \citep{mackinlay1997event} and \citep{campbell1997econometrics}. Normal performance is measured with a constant mean return model\footnote{A constant mean return model is used instead of a market model in order to maximize the number of observations (plausible market indices such as the MSCI World Index and the S\&P/IFC Emerging Markets Investable Composite Index only begin in 1976 and 1995 respectively). That said, results are insensitive to the use of a market model.},
\begin{align} \label{eqn:market-model}
R_{it}=\mu_{i}+\epsilon_{it},
\end{align}
where $R_{it}$ is the logged return of national stock index $i$ on trading day $t$ and $\epsilon_{it}$ is the error term. We calculate abnormal returns (ARs), in an ``event window'' surrounding the date of each coup, $AR_{i\tau}=R_{i\tau}-\widehat{\mu}_i$, where $\tau$ is a date in the event window, and $\widehat{\mu}_i$ is estimated in an ``estimation window'' preceding the event window with \autoref{eqn:market-model}. We use a 41 day event window (i.e. 20 pre-event trading days, the event day, and 20 post-event trading days). The estimation window is the 250 trading days prior to the start of the event window. The abnormal returns are then used to generate cumulative abnormal returns (CARs) between event day $\tau_1$ and event day $\tau_2$: $CAR(\tau_1,\tau_2)=\sum_{\tau=\tau_1}^{\tau_2}AR_{i\tau}$.

@@

We report abnormal returns separately for coups, assassinations and resignations because they may have distinct effects on stock returns. Standard errors and p-values are calculated using asymptotic t-statistics as in \citep{mackinlay1997event}.\footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).}

@@

\subsection{Volatility} \label{subsec: Volatility}
Although irregular regime changes have disparate effects on the direction of stock returns depending on expected policy outcomes, the results from \autoref{subsec: Abnormal Returns} suggest that all of these political events increase financial volatility. However, since stock volatility is not directly observable, one must decide how to best estimate volatility. Our estimates are obtained from a generalized autoregressive conditional heteroskedasticity (GARCH) model estimated using 1000 pre-event days, the event day and 1000 post-event days. As in \citep{jensen2005market} and \citep{leblang2005government}, we use the GARCH (1,1) specification. In particular, for national stock index $i$,
\begin{align*}
R_{it}=\mu_i + \epsilon_{it},\hspace{1cm} \epsilon_{it}\sim \mathcal{N}\left(0,\sigma_{it}^2\right),
\end{align*}
where $\mu_i$ is a constant and,
\begin{align*}
\sigma_{it}^2&=\gamma_{i}+\alpha_{i}\epsilon_{i,t-1}^2+\beta_{i}\sigma_{i,t-1}^2.
\end{align*}
The key parameter of interest is the conditional variance, $\sigma_{it}^2$. The one-period-ahead volatility forecasts, $\sigma_{it}$, are larger when $\epsilon_{i,t-1}^2$ and $\sigma_{i,t-1}^2$ are larger. In other words, the model predicts that large shocks will be followed by other large shocks.

@@

To explore the first concern, we create a synthetic control portfolio for each event based on the techniques introduced in \citep{abadie2003economic} and \citep{abadie2010synthetic}. Each country is given a weight which represents its influence in the synthetic control portfolio. The weight is chosen so that the daily returns and the variance of the daily returns of the control portfolio and the event country are most similar in the estimation window. The set of possible countries in the control portfolio consists of all countries listed in \autoref{tab:stock list}.\footnote{See the appendix for a more formal explanation.}

The second concern is addressed using non-parametric statistical techniques, which are free from distributional assumptions. We employ the sign and the rank tests which are based on the sign and the rank of the event day ARs respectively.\footnote{See section 8 in \citep{mackinlay1997event} for more details.} Both tests are less influenced by departures from normality than statistics based on traditional t-tests such as those reported earlier in this paper.

@@

\subsection{Synthetic Control Portfolio}
Let $\boldsymbol{R}_{k}$ be the vector of returns for the event country in the estimation window, $\boldsymbol{R}_{-k}$ be the vector of returns for all other countries in the estimation window, $\boldsymbol{X}_1=(\boldsymbol{R}_{k},\var(\boldsymbol{R}_{k}))$, $\boldsymbol{X}_0=(\boldsymbol{R}_{-k},\var(\boldsymbol{R}_{-k}))$, and $\boldsymbol{W}_{-k}$ be a $((N-1) \times 1)$ vector of weights where $N$ is the number of countries listed in \autoref{tab:stock list}. Then $\boldsymbol{W}^*$ is chosen to minimize $(\boldsymbol{X}_1-\boldsymbol{X}_0\boldsymbol{W})'\boldsymbol{V}(\boldsymbol{X}_1-\boldsymbol{X}_0\boldsymbol{W})$ subject to $w_i\geq0$ $(i = 1,2,\ldots,N-1)$ and $\sum_i^{N-1} w_i = 1$, and the vector $\boldsymbol{V}$ is chosen so that stock returns for the control portfolio during the estimation window are are close as possible to the event country.\footnote{See \citep{abadie2003economic} for further details.}

>>>>

A common hypothesis is that political instability discourages investment and reduces economic growth by increasing policy uncertainty. This hypothesis is supported by cross country empirical evidence that political instability - as measured by regime change frequency or political violence - is negatively correlated with investment, GDP growth, and financial development in cross country regressions \citep{alesina1996political,alesina1996income,roe2011political}. \footnote{In particular, \citet{alesina1996political} show that countries have lower growth in years in which there are major government changes such as coups. On the other hand, \citet{londregan1990poverty} finds that coups have no effect on economic growth.} However, such hypotheses are overly simplistic. Not all political instability is equivalent - resignations, assassinations, and coups are each examples of political instability, but may have disparate effects on financial returns. Further, political instability may be seen as a \textit{positive} event if the current regime is regarded as strongly anti-business or anti-global.  \citet{alesina1996political} takes the possibility of a new government adopting better economic policies seriously, but argues that the negative effects of uncertainty dominate the positive effects of coups staged by pro growth factions.

As in \citet{alesina1996political}, we use ``irregular'' regime changes such as coups, assassinations and resignations as an indicator for political instability. The major innovation of this paper is that, in contrast to the aforementioned cross country studies, we analyze the impact of political instability using daily financial data. This enables us to use market expectations to quantify the economic impact of each individual type of regime change. The advantage of this event study approach is that it allows us to isolate the effect of irregular regime changes and mitigate the endogeneity problems that are nearly ubiquitous in cross country regressions.

@@

This paper is closely related to four branches of literature. The first analyzes the economic impact of political instability. This area contains a number of theoretical models in addition to the empirical work already mentioned. These models tend to focus on government policy instruments that affect private rates of return on investment. The general argument is that political instability causes governments to act myopically and adopt inefficient policies.\footnote{For instance, \citet{svensson1998investment} argues that governments are less likely to invest in the legal system and the protection of property rights when political uncertainty is high. Similarly, \citet{devereux1998political} propose a model in which political instability causes governments to leave fewer assets to their successors which forces them into increasing capital taxes. The knowledge of future taxation causes the private sector to reduce current investment which reduces future output.}

The second looks at how the variance of stock market returns reacts to political events. \citet{leblang2005government} show that right-wing governments increase the mean and volatility of stock prices in the United States and Britain. \citet{jensen2005market} find that a rise in the popularity of a Brazilian presidential candidate with uncertain future policies increased the volatility (but not the mean return) of the Brazilian stock market.

The third strand examines the impact of leaders on political and economic outcomes. \citet{jones2005leaders} identify exogenous leadership transitions using deaths of leaders from natural causes or accidents. Their results indicate that leaders influence monetary policy and economic growth. \citet{jones2009hit} find that transitions to democracy are more likely after the assassinations of autocrats.

@@

\section{Impact of Political Instability on Stock Returns} \label{sec: Impact of Political Instability on Stock Returns}
\subsection{Abnormal Returns} \label{subsec: Abnormal Returns}
We begin our analysis by studying the effect of irregular regime changes on stock returns. We follow the standard event study methodology as presented by, among others, \citet{mackinlay1997event} and \citet{campbell1997econometrics}. Normal performance is measured with a constant mean return model\footnote{A constant mean return model is used instead of a market model in order to maximize the number of observations (plausible market indices such as the MSCI World Index and the S\&P/IFC Emerging Markets Investable Composite Index only begin in 1976 and 1995 respectively). That said, results are insensitive to the use of a market model.},
\begin{align} \label{eqn:market-model}
R_{it}=\mu_{i}+\epsilon_{it},
\end{align}
where $R_{it}$ is the logged return of national stock index $i$ on trading day $t$ and $\epsilon_{it}$ is the error term. We calculate abnormal returns (ARs), in an ``event window'' surrounding the date of each coup, $AR_{i\tau}=R_{i\tau}-\widehat{\mu}_i$, where $\tau$ is a date in the event window, and $\widehat{\mu}_i$ is estimated in an ``estimation window'' preceding the event window with \autoref{eqn:market-model}. We use a 41 day event window (i.e. 20 pre-event trading days, the event day, and 20 post-event trading days). The estimation window is the 250 trading days prior to the start of the event window. The abnormal returns are then used to generate cumulative abnormal returns (CARs) between event day $\tau_1$ and event day $\tau_2$: $CAR(\tau_1,\tau_2)=\sum_{\tau=\tau_1}^{\tau_2}AR_{i\tau}$.

@@

We report abnormal returns separately for coups, assassinations and resignations because they may have distinct effects on stock returns. Standard errors and p-values are calculated using asymptotic t-statistics as in \citet{mackinlay1997event}.\footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).}

@@

\subsection{Volatility} \label{subsec: Volatility}
Although irregular regime changes have disparate effects on the direction of stock returns depending on expected policy outcomes, the results from \autoref{subsec: Abnormal Returns} suggest that all of these political events increase financial volatility. However, since stock volatility is not directly observable, one must decide how to best estimate volatility. Our estimates are obtained from a generalized autoregressive conditional heteroskedasticity (GARCH) model estimated using 1000 pre-event days, the event day and 1000 post-event days. As in \citet{jensen2005market} and \citet{leblang2005government}, we use the GARCH (1,1) specification. In particular, for national stock index $i$,
\begin{align*}
R_{it}=\mu_i + \epsilon_{it},\hspace{1cm} \epsilon_{it}\sim \mathcal{N}\left(0,\sigma_{it}^2\right),
\end{align*}
where $\mu_i$ is a constant and,
\begin{align*}
\sigma_{it}^2&=\gamma_{i}+\alpha_{i}\epsilon_{i,t-1}^2+\beta_{i}\sigma_{i,t-1}^2.
\end{align*}
The key parameter of interest is the conditional variance, $\sigma_{it}^2$. The one-period-ahead volatility forecasts, $\sigma_{it}$, are larger when $\epsilon_{i,t-1}^2$ and $\sigma_{i,t-1}^2$ are larger. In other words, the model predicts that large shocks will be followed by other large shocks.

@@

To explore the first concern, we create a synthetic control portfolio for each event based on the techniques introduced in \citet{abadie2003economic} and \citet{abadie2010synthetic}. Each country is given a weight which represents its influence in the synthetic control portfolio. The weight is chosen so that the daily returns and the variance of the daily returns of the control portfolio and the event country are most similar in the estimation window. The set of possible countries in the control portfolio consists of all countries listed in \autoref{tab:stock list}.\footnote{See the appendix for a more formal explanation.}

The second concern is addressed using non-parametric statistical techniques, which are free from distributional assumptions. We employ the sign and the rank tests which are based on the sign and the rank of the event day ARs respectively.\footnote{See section 8 in \citet{mackinlay1997event} for more details.} Both tests are less influenced by departures from normality than statistics based on traditional t-tests such as those reported earlier in this paper.

@@

\subsection{Synthetic Control Portfolio}
Let $\boldsymbol{R}_{k}$ be the vector of returns for the event country in the estimation window, $\boldsymbol{R}_{-k}$ be the vector of returns for all other countries in the estimation window, $\boldsymbol{X}_1=(\boldsymbol{R}_{k},\var(\boldsymbol{R}_{k}))$, $\boldsymbol{X}_0=(\boldsymbol{R}_{-k},\var(\boldsymbol{R}_{-k}))$, and $\boldsymbol{W}_{-k}$ be a $((N-1) \times 1)$ vector of weights where $N$ is the number of countries listed in \autoref{tab:stock list}. Then $\boldsymbol{W}^*$ is chosen to minimize $(\boldsymbol{X}_1-\boldsymbol{X}_0\boldsymbol{W})'\boldsymbol{V}(\boldsymbol{X}_1-\boldsymbol{X}_0\boldsymbol{W})$ subject to $w_i\geq0$ $(i = 1,2,\ldots,N-1)$ and $\sum_i^{N-1} w_i = 1$, and the vector $\boldsymbol{V}$ is chosen so that stock returns for the control portfolio during the estimation window are are close as possible to the event country.\footnote{See \citet{abadie2003economic} for further details.}
